\chapter{Project proposal} % (fold)
\label{cha:project_proposal}

In the Part \ref{part:overview_of_the_topic} we have presented the current state of the research of the P system variants. Especially the parallelism options have been investigated very well, but there are still some gaps that should be filled. For many variants the universality has been proven only when working in the maximally parallel mode. In most cases the sequential mode is strictly weaker, but this is not always true.

We have studied several variants of sequential P systems in order to obtain universality without using maximal parallelism.

\section{Sequential P systems without priorities with cooperative rules} % (fold)
\label{sec:sequential_p_systems_without_priorities_with_cooperative_rules}

A sequential variant without priorities and with cooperative rules is not univesal (see \cite{Ibarra04dang}). They have tried modifying the variant to make it universal and showed that with rules for membrane creation with unbounded number of membranes it became universal.

We tried other approach using rules with inhibitors. This variant is universal in both generating and accepting case and we don't even have to use membrane dissolution rules. In generative case the model is able to simulate maximal parallel P system and the accepting model can simulate a register machine.

Future plans include research other more restricted variants such as omitting cooperation in the rules or restrict the power of inhibitors.

\section{P systems with vacuum} % (fold)
\label{sec:p_systems_with_vacuum}

We propose a new variant of P system with vacuum in accordance with the criteria formulated in \cite{Besozzi:PhD:2004} (see beginning of the section \ref{sec:p_system_variants}). In the common sense, vacuum represents a state of space with no or a little matter in it. Using vacuum in modelling frameworks can help express certain phenomena more easily. We define a new P system variant, which creates a special vacuum object in a region as soon as the region becomes empty. The vacuum is removed whenever some object interacts with it. After the interaction, there is vacuum no longer. This removal process is realized by allowing the vacuum object to be used only on the left side of rules. If we made the vacuum to be removed automatically when an object enters the region, there would be no difference with the variant without vacuum objects because of no interactions with it.

We are interested in how the variant with the vacuum improves the computation power of a sequential P system in comparison to the variant without using the vacuum. This case have been shown to be universal.

In the future we will research other restriction for this variant such as non-cooperative rules, decaying objects, deterministic steps and will use techniques to show even non-universality results.

% section p_systems_with_vacuum (end)

% chapter project_proposal (end)

\chapter{Current status of the research} % (fold)
\label{cha:current_status_of_the_research}

\section{Sequential P systems without priorities with cooperative rules and inhibitors} % (fold)
\label{sec:sequential_p_systems_without_priorities_with_cooperative_rules_and_inhibitors}

Original definition of P systems with inhibitors (see \cite{Ionescu:jucs_10_5:on_p_systems_with}) allow to use only one inhibitor pre rule, e.g. $u\rightarrow v|_{\neg i}$. Alternative definition (see \cite{Agrigoroaiei:2010:Dissolution}) allow to use whole inhibitor set in the rule like $u\rightarrow v|_{\neg B}$, where $B$ is a set of objects. Such a rule can be applied only if no element of $B$ is present in the region.

For sequential P systems, the following lemma will show the equivalence of these definitions. However, special condition must be fulfilled, the regions with such rules cannot be empty.

\input{seqinhproof}

We have also reached this result in the accepting case by simulation of register machines.

\input{../inh/seqinhaccproof}

% section sequential_p_systems_without_priorities_with_cooperative_rules_and_inhibitors (end)

\section{Universality results of P systems with vacuum} % (fold)
\label{sec:universality_results_of_p_systems_with_vacuum}

\input{../inh/vacuumproof}

% section universality_results_of_p_systems_with_vacuum (end)

% chapter current_status_of_the_research (end)

