% !TEX root = diz.tex
\chapter*{Acknowledgments}

I am very grateful to all the people I had the privilege to work with, especially I would like to thank my advisor RNDr. Damas Gruska PhD. for his constant interest, encouragement and help that I deeply appreciate.
The present work was possible due to many beneficial consultations and intensive cooperation.

The major thanks also goes to those who provided valuable feedback to my work, specifically classmates Mgr. Martin Králik and Mgr. Andrej Borsuk and the editors and referees of conferences I has contributed to: Hybrid Systems and Biology 2013, Computational Models for Cell Processes 2013 and Computability in Europe 2014, 2015.

I would also like to thank to the members of Laboratory of Comparative and Functional Genomics of Eukaryotic Organelles and their head, Prof. Ľubomír Tomáška for familiarization with the work of biologists.

Last, but not least, I thank to my family, colleagues, teammates and all the friends who supported me especially last weeks before the deadlines for their patience, support and understanding.
\ifdefined\godzilla
  Especially for their understanding that they will probably never understand my work.

  Of course I cannot forget Branislav Jobus and his book \cite{Jobus10Ancijas} which gave me new motivation and inspired me to continue this work. 
\fi

% Osobitná vďaka patrí vedúcemu diplomovej práce doc. RNDr. Damasovi Gruskovi PhD. za cenné rady, námety, podnetné pripomienky a všestrannú pomoc, ktorú si hlboko vážim. Len vďaka mnohým prínosným konzultáciam a intenzívnej spolupráci som bol schopný napísať toto dielo. Nesmiem zabudnúť ani na RNDr. Branislava Rovana, CSc. a spolužiakov za to, že si na dizertačnom seminári našli čas, aby si vypočuli moju prezentáciu dizertačnej práce. Ďalšie poďakovania venujem rodičom a známym, ktorí to so mnou dokázali vydržať posledné týždne pred odovzdaním.
