% !TEX root = ../diz.tex
\index{P systems!sequential} A sequential variant without priorities and with cooperative rules is not universal (see \cite{Ibarra04dang}). They have tried modifying the variant to increase the computational power and showed that with rules for membrane creation with unbounded number of membranes it became universal.

We have tried another approach using rules with inhibitors. We show that this variant is computationally complete in both generating and accepting case. For the generative case we present a proof by a simulation of maximal parallel P system and in the accepting case we prove it can simulate a register machine. These results were published in \cite{Kovac14Inhibitors}.

In the literature there are two variants of \index{P systems!with inhibitors} rules with inhibitors.
\begin{enumerate}
  \item Some of them (\cite{Ionescu:jucs_10_5:on_p_systems_with}, \cite{Sburlan05dragos}) allow to use only one inhibitor per rule, e.g. $u\rightarrow v|_{\neg i}$.
  \item Others (\cite{Agrigoroaiei:2010:Dissolution}, \cite{Sburlan:2006:FurtherResultsPromotersInhibitors}) are more expressive allowing to use a set of inhibitors per rule, e.g. $u\rightarrow v|_{\neg B}$, where $B$ is a set of objects. Such a rule can be applied only if no element of $B$ is present in the region, where the rule is applied.
\end{enumerate}

Our proof uses variant of rules with a set of inhibitors, but they could also be implemented with rules with a single inhibitor, while the complexity of the simulation would be increased.
Dissolution could be also solved with increased complexity by introducing additional phases of the simulation. But because P systems without dissolution are still computationally complete \cite{Agrigoroaiei:2010:Dissolution}, and for the sake of simplicity, we simulate only rules without dissolution 

\begin{definition}
\label{def:evolution_rule_without_dissolution}
  An {\bf evolution rule without dissolution} over an alphabet $\Sigma$ in a P system with $m$ membranes is a pair $(u,v)$, which is usually written as $u\rightarrow v$, where $u$ is a string over $\Sigma$ and $v$ is a string over $\Sigma\times(\{here, out\}\cup\{in_j|1\leq j\leq m\})$.
\end{definition}

\begin{definition}
  An {\bf evolution rule with inhibitors without dissolution} is a triple $(u,v,B)$, which is usually written as $u\rightarrow v|_{\neg B}$, where $(u,v)$ is an evolution rule without dissolution by definition \ref{def:evolution_rule_without_dissolution} and $B\subseteq\Sigma$ is a set of objects called inhibitors of the rule.
\end{definition}

\input{seqinhproof}

We have also reached this result in the accepting case by simulation of \index{Register machine} register machines.

\input{seqinhaccproof}

\subsection{Concluding remarks} % (fold)
\label{sub:concluding_remarks_of_inhibitors}

Although these results are not very surprising as similar results already hold for Petri nets, which are equivalent to sequential P systems, the simulation of maximal parallel P system used in the proof for generating case is valuable not only for the universality, but also can be seen as a method of conversion between P systems in sequential manner and maximally parallel manner, which may be essential for future works on P systems and other multiset rewriting systems. The simulation also shows a method how to synchronize application of multiple rules in a membrane and how to synchronize this parallel rule application across whole membrane structure. Sequential variants are promising alternative to traditional maximal parallel variants and will be good subject for the further research. Future plans include research of other more restricted variants such as omitting cooperation in the rules or restricting the power of inhibitors.

% subsection concluding_remarks (end)
