% !TEX root = ../diz.tex
In subsection \ref{sub:membrane_systems} we briefly introduced the notions of membrane, membrane structure and P system. Chapter \ref{cha:preliminaries} introduced some basic definitions useful for better understanding of further reading. In this Chapter we define a P system (section \ref{sec:definitions}), its variants with recent advances (section \ref{sec:p_system_variants}) and case studies (section \ref{sec:case_studies}).

A P system \cite{Paun98} is a computing model which abstracts from the way the alive cells process chemical compounds in their compartmental structure. In short, in the regions defined by a membrane structure we have objects which can be described by symbols. Their multiplicity matters, that is, we work with multisets of objects placed in the regions of the membrane structure. Objects evolve according to given rules. Any object, alone or together with another objects, can be transformed in other objects, can pass through a membrane, and can dissolve the membrane in which it is placed. All objects evolve at the same time, in parallel manner across all membranes. The evolution rules are hierarchized by a priority relation, which is a partial order. By using the rules in a nondeterministic, maximally parallel manner, one gets transitions between the system configurations. A sequence of transitions is a computation. With a halting computation we can associate a result, in the form of the objects present in a given membrane in the halting configuration, or expelled from the system during the computation.

% P system

These aspects all together forms a P system as introduced in \cite{Paun98}. An example of a P system containing rules for computing the Fibonacci sequence can be seen in the figure \ref{fig:p_system_fibonacci}.

In section~\ref{sec:definitions} we will provide a formal definition of a P system. We will follow in section \ref{sec:p_system_variants} by presenting most common P system variants along with a survey of known results and in section \ref{sec:case_studies} we show how P systems can be of use in practice case studies.

\section{Definitions} % (fold)
\label{sec:definitions}
\input{p-system-definition}

% section definitions (end)

\section{P system variants} % (fold)
\label{sec:p_system_variants}
\input{variants}

% section p_system_variants (end)

\section{Case studies} % (fold)
\label{sec:case_studies}

\input{casestudies}

% section case_studies (end)