\subsection{Context rules instead of cooperative rules}
  In the proof, we use cooperative rules, but for the convenience, we will write more than two objects at the left side of some rules.
  \begin{lemma}
  \label{lemma:context_rules}
    For sequential P system, the variant with cooperative rules and the variant with context rules are equal if at most one reactant in the context rule is consumed in all other rewrite rules.
  \end{lemma}
  \begin{dokaz}
    Consider rule $a_1|a_2|\dots|a_n \rightarrow v$, where $a_1,\dots a_n$ are objects and $v$ is a product of the rule, which may include sending object up/down through the membrane.
    This rule can be simulated with multiple sequential steps:
    \begin{itemize}
      \item $a_1|a_2 \rightarrow a_{1,2}$
      \item $a_{1,2}|a_3 \rightarrow a_{1,2,3}$
      \item \dots
      \item $a_{1,2,\dots n-1}|a_n \rightarrow v$
    \end{itemize}    
  \end{dokaz}
  There could be a problem with a case when these multiple sequential steps would be interrupted in the middle and the rewriting would be left incomplete. Therefore, we have the additional condition of having at most one reactant being used and consumed in other rule. That reactant will be referred as $a_1$. So after the first step is executed, there is no way to be interrupted.

\subsection{Inhibitor set}
Original definition of P system with inhibitors (see \cite{Ionescu:jucs_10_5:on_p_systems_with}) have only a single inhibitor object per rule. P systems with inhibitor set can have rules of type $u\rightarrow v|_{\neg B}$, where $B\subseteq V$. The rule $u\rightarrow v$ is active only if there is no occurrence of any symbol from $B$.

Lemma \ref{lemma:inhibitor_step} will establish equivalence of these two definitions of rewriting rule when special precondition is fulfilled - there is no empty region.

\begin{lemma}
\label{lemma:inhibitor_step}
  If there is at least one object present in each region of a P system, rewriting step in P system with inhibitor set can be simulated by multiple consecutive steps of P system with single inhibitor.
\end{lemma}

\begin{dokaz}
  Consider P system with alphabet $V$ and a set of inhibitors $B=\{b_1, b_2, \dots ,b_n\}$.
  We introduce new symbols $GONE_{b_1}, GONE_{b_2}, \dots , GONE_{b_n}$, where the presence of $GONE_{b_i}$ in a membrane means that there is no occurence of $b_i$.
  For each rule that use the inhibitor set $B$ such as $u\rightarrow v|_{\neg B}$ we will have rules:
  \begin{itemize}
    \item $c \rightarrow c|GONE_{b}|_{\neg b}$ for all $ c\in V, b\in B$
    \item $u|GONE_{b_1}|GONE_{b_2}|\dots|GONE_{b_n} \rightarrow v|GONE_{b_1}|GONE_{b_2}|\dots|GONE_{b_n}$
  \end{itemize}
\end{dokaz}
